---
title: "D-score versus Logit-scale"
author: "Stef van Buuren"
date: "2024-08-02"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 6, fig.height = 6)
# library(dmetric)
```

## Background

We look into differences between the Logit and D-score scales. The D-score is a linear transformation of the Logit scale. We therefore expect that

- D-score and Logit are linearly related, so should lay on a line
- SEM(D-score) and SEM(Logit) are proportional

## Milestones data

```{r milestones}
library(dscore)

data <- milestones
dim(data)
```

## Calculate both metrics - MILESTONES

```{r}
# D-score
ds <- dscore(data = data, xname = "agedays", xunit = "days")
head(ds)

# Logit
dl <- dscore(data = data, xname = "agedays", xunit = "days", metric = "logit")
head(dl)
```

## Control plots - MILESTONES

```{r}
# Plots
betas <- builtin_keys[builtin_keys$key == "gsed2406", c("intercept", "slope")]
plot(x = dl$d, y = ds$d, cex = 0.4, pch = 19,
     xlab = "D-score (logit)", ylab = "D-score (dscore)",
     main = "D-score (logit) vs D-score (dscore) MILESTONES")
abline(coef = betas, col = "orange")
plot(x = dl$sem, y = ds$sem, cex = 0.4, pch = 19,
     xlab = "SEM (logit)", ylab = "SEM (dscore)",
     main = "SEM (logit) vs SEM (dscore) MILESTONES")
abline(0, betas[2], col = "orange")
```

## 293_0 Data

```{r}
path <- file.path("~/Project/gsed/phase1/202407/293_0")

model <- readRDS(file.path(path, "model.Rds"))
data <- readRDS(file.path(path, "data.Rds"))
dim(data)
```

## Calculate both metrics - 293_0

```{r}
# D-score
ds <- dscore(data = data, xname = "agedays", xunit = "days")
head(ds)

# Logit
dl <- dscore(data = data, xname = "agedays", xunit = "days", metric = "logit")
head(dl)
```

## Control plots - 293_0

```{r}
# Plots
betas <- builtin_keys[builtin_keys$key == "gsed2406", c("intercept", "slope")]
plot(x = dl$d, y = ds$d, cex = 0.4, pch = 19,
     xlab = "D-score (logit)", ylab = "D-score (dscore)",
     main = "D-score (logit) vs D-score (dscore) 293_0")
abline(coef = betas, col = "orange")
plot(x = dl$sem, y = ds$sem, cex = 0.4, pch = 19,
     xlab = "SEM (logit)", ylab = "SEM (dscore)",
     main = "SEM (logit) vs SEM (dscore) 293_0")
abline(0, betas[2], col = "orange")
```

## Full data

```{r model818_6}
library(dmetric)

# get model and data
model_name <- "818_6"
path <- file.path("~/Project/gsed/phase1/202407", model_name)
model <- readRDS(file.path(path, "model.Rds"))
data <- readRDS(file.path(path, "data.Rds"))
dim(data)
```

## Calculate both metrics - FULL

```{r}
# D-score
ds <- dscore(data = data, xname = "agedays", xunit = "days")
head(ds)

# Logit
dl <- dscore(data = data, xname = "agedays", xunit = "days", metric = "logit")
head(dl)
```

## Control plots - FULL

```{r}
# Plots
betas <- builtin_keys[builtin_keys$key == "gsed2406", c("intercept", "slope")]
plot(x = dl$d, y = ds$d, cex = 0.4, pch = 19,
     xlab = "D-score (logit)", ylab = "D-score (dscore)",
     main = "D-score (logit) vs D-score (dscore) FULL")
abline(coef = betas, col = "orange")
plot(x = dl$sem, y = ds$sem, cex = 0.4, pch = 19,
     xlab = "SEM (logit)", ylab = "SEM (dscore)",
     main = "SEM (logit) vs SEM (dscore) FULL")
abline(0, betas[2], col = "orange")
```

## Control plots - FULL - by study

```{r}
library(dplyr)
library(ggplot2)
plotdata <- cbind(ds, logitd = dl$d, logitsem = dl$sem, study = data$cohort) 
plotdata <- filter(plotdata, study %in% c("GSED-BGD", "GSED-PAK", "GSED-TZA"))
g1 <- ggplot(data = plotdata, aes(x = logitd, y = d, color = study)) +
  theme_minimal() +
  geom_point(size = 0.1) +
  # geom_abline(intercept = betas[1], slope = betas[2], color = "orange") +
  labs(y = "D-score", x = "Logit") +
  facet_wrap(~study) +
  theme(legend.position = "none")
g1
```

```{r}
g2 <- ggplot(data = plotdata, aes(x = logitsem, y = sem, color = study)) +
  theme_minimal() +
  geom_point(size = 0.1) +
  # geom_abline(intercept = betas[1], slope = betas[2], color = "orange") +
  labs(y = "D-score", x = "Logit") +
  facet_wrap(~study) +
  theme(legend.position = "none")
g2
```


